% !TEX root=main.tex

\chapter{THE ROMAN EMPIRE}

\linenumbers

Rome is in Italy. Italy is in Europe. Greece is in Europe. Italy and Rome are in Europe. Spain is also in Europe. Spain and Italy and Greece are in Europe.

Egypt is not in Europe, Egypt is in Africa. Gaul is not in Africa, Gaul is in Europe. Syria is not in Europe, but in Asia. Arabia is also in Asia. Syria and Arabia are in Asia. Germany is not in Asia, but in Europe. Britain is also in Europe. Germany and Britain are in Europe.

Is Gaul in Europe? Gaul is in Europe. Is Rome in Gaul? Rome is not in Gaul. Where is Rome? Rome is in Italy. Where is Italy? Italy is in Europe. Where are Gaul and Spain? Gaul and Spain are in Europe.

Is the Nile in Europe? The Nile is not in Europe. Where is the Nile? The Nile is in Africa. Where is the Rhine? The Rhine is in Germany. The Nile is a river. The Rhine is a River. The Nile and the Rhine are rivers. The Danube is also a river. The Rhine and the Danube are rivers in Germany. The Tiber is a river in Italy.

The Nile is a large\footnote{\emph{magnus} v. \emph{parvus}} river. The Tiber is not a large river, the Tiber is a small river. The Rhine is not a small river, but a large river/one*. The Nile and the Rhine are not small rivers, but large rivers/ones*. The Danube is also a large river.

Corsica is an island. Corsica and Sicily are islands. Britain is also an island. Italia is not an island. Sicily is a large island. Melita is a small island. Britain is not a small island, but a large one. Sicily and Sardinia are not small islands, but large ones.

Brindisi is a town. Brindisi and Tusculum are towns. Sparta is also a town. Brindisi is a large town. Tusculum is a small town. Delphi is also a small town. Tusculum and Delphi are not large towns, but small ones.

Where is Sparta? Sparta is in Greece. Sparta is a Greek town. Sparta and Delphi are Greek towns. Tusculum is not a Greek city, but a Roman one. Tusculum and Brindisi are Roman towns. Sardinia is a Roman island. Crete, Rhodes, Naxos, Samos, Chios, Lesbos, Lemnos and Euboea are Greek islands. There are lots of islands in Greece. There are lots of towns in Italy and Greece. There are many rivers in Gaul and Germany. Are there many rivers and many towns in Arabia? There are not many, but few rivers and few towns.

Is Crete a town? Crete is not a town! What is Crete? Crete is an island. Is Sparta an island? Sparta is not an island! What is Sparta? Sparta is a town. What is the Rhine? The Rhine is a large river. Is the Atlantic ocean small? It is not a small, but a large ocean.

Where is the Roman empire? The Roman empire is in Europe, in Asia and in Africa. Spain, Syria and Egypt are Roman provinces. Germany is not a Roman province: Germany is not in the Roman empire. Gaul and Britain are also Roman provinces. There are many provinces in the Roman empire. The Roman empire is large!

\section[Letters and numbers]{LETTERS AND NUMBERS}
{\sc i} and {\sc ii} are numbers. {\sc iii} is also a number. {\sc i}, {\sc ii} and {\sc iii} are Roman numbers. {\sc i} and {\sc ii} are small numbers. {\sc ci)**} is a large number.

{\sc a} and {\sc b} are letters. {\sc c} is also a letter. {\sc a}, {\sc b} and {\sc c} are three letters. {\sc a} is the first letter ({\sc i}), {\sc b} is the second letter ({\sc ii}) and {\sc c} is the third letter ({\sc iii}). {\sc γ} is a Greek letter. {\sc c} is a Latin letter. {\sc c} and {\sc d} are Latin letters. {\sc γ} and {\sc δ} are Greek letters.

\emph{Fluvius} and \emph{oppidum} are Latin words. \emph{Ubi} is also a Latin word. The word \emph{ubi} has\footnote{Literally, `in the word \emph{ubi} there are'.} three letters. In the first chapter there are a thousand words.

The word \emph{insula} has six letters and three syllables: the first syllable is \emph{in-}, the second \emph{-su-} and the third \emph{-la}. The word \emph{non} has three letters and one syllable.

What is {\sc iii}? {\sc iii} is a Roman number. What is {\sc γ}? \smallc γ is a Greek letter. Is {\sc c} a Greek letter? {\sc c} is not a Greek letter, but a Latin one. Is {\sc b} the first letter? {\sc b} is not the first letter, but the second one. What is \emph{non}? \emph{Non} is a Latin word. \emph{Non, sed, magnus} and \emph{numerus} are Latin words. \emph{Vocabulum} is also a Latin word!\footnote{Word play; \emph{vocabulum} means `word'.}

\section[Grammar]{LATIN GRAMMAR}
\subsection{\emph{Singular and plural}}

\begin{enumerate}[(A)]
\item \emph{-us} --> \emph{-i}\\
  \emph{Fluvi}us -> \emph{Fluvi}i
\item \emph{-a} --> \emph{-ae}\\
  \emph{insul}a --> \emph{insul}ae
\item \emph{-um} --> \emph{-a}\\
  \emph{oppid}um --> \emph{oppid}a
\end{enumerate}

\nolinenumbers

\section[Exercises]{EXERCISES}
\subsection*{EXERCISE A}
Nilus fluvius est. Nilus et Rhenius fluvii sunt. Creta insula est. Creta et Rhodus insulae sunt. Brundisium oppidum est. Brundisium et Tusculum oppida sunt.

Rhenus fluvius magnus est. Tiberis est fluvius parvus. Rhenus et Danuvius non fluvii parvii, sed fluvii magni sunt. Sardinia insula magna est. Melita insula parva est. Sardinia et Sicilia non insulae parvae, sed insulae magnae sunt. Brundisium non oppidum parvum, sed oppidum magnum est. Tusculum et Delphi non oppida magna, sed oppida parva sunt.

Creta insula Graeca est. Lesbos et Chios et Naxus sunt insulae Graecae. In Graecia multae insulae sunt. In Gallia sunt multii fluvii. In Italia mula oppida sunt. In Arabia sunt paucii fluvii et pauca oppida.

\smallc a et \smallc b litterae Latinae sunt. \smallc c quoque littera Latina est. \emph{Multi} et \emph{pauci} vocabula Latina sunt. \emph{Ubi} quoque vocabulum Latinum est. \smallc i et \smallc{ii} numeri Romani sunt. \smallc{iii} quoque numerus Romanus est.

\subsection*{EXERCISE B}
Sicilia in Italia est. Italia insula non est. Rhenus fluvius est. Brundisium oppidum est. Sicilia est Sardinia insulae magnae sunt. Melita insula...

\subsection*{EXERCISE C}
...
