% !TEX root=main.tex

\chapter{THE RUDE BOY}

\linenumbers[1]

\section[First scene]{FIRST SCENE}
\emph{Characters: Julia, Marcus, Quintus.}

Julia sings: ``Lalla.'' Julia is cheerful.

Marcus: ``St!'' Marcus is not cheerful.

Julia sings: ``Lalla, lalla.''

Marcus: ``Ssst!'' Marcus is angry.

Julia sings: ``Lalla, lalla, lalla.''

Marcus beats Julia.

Julia doesn't sing anymore, but she cries: ``Uhuhu!''

Marcus laughs: ``Hahahae!''

Quintus sees Marcus. Marcus doesn't see Quintus.

Quintus: ``What? Marcus beats the girl--and laughs!''

Quintus is angry and beats Marcus! Marcus doesn't laugh anymore. Marcus, angry, beats Quintum.

Julia: ``Where is mother?'' Julia doesn't see Aemilia.

Julia calls mother. ``Mother! Marcus has beat Quintus!''

Marcus (angry): ``St!'' Marcus beats Julia.

Julia cries and calls Aemilia: ``Mom! Mom! Marcus has beat me!''

Aemilia comes.

\section[Second scene]{SECOND SCENE}
\emph{Characters: Aemilia, Julia, Marcus, Quintus.}

Aemilia asks: ``Who calls me?''

Quintus responds: ``Julia calls you.''

Aemilia asks Quintus: ``Why does Julia cry?''

Quintus responds: ``Julia cries, because Marcus has beat her.''

Aemilia: ``What? The boy has hit the little girl? Fu*** Why has Marcus beat Julia?''

Quintus: ``Because Julia sings.''

Aemilia: ``Oh Julia, my little daughter! Marcus is not a good boy; Marcus is a rude boy!''

Quintus: ``Julia is a good girl.''

Aemilia asks Quintus: ``Where is Julius? Why doesn't he come?'' Aemilia doesn't see Julius.

Marcus responds: ``Father sleeps.''

Quintus: ``Mother asks me, not you!''

Aemilia: ``St, boys! Where is father?''

Quintus: ``Father is not here, but Marcus is.''

Quintus calls Julius: ``Father! Fa-ther!''

Julius doesn't hear Quintus and doesn't come. Why does Julius not hear Quintus? Julius doesn't hear him because he sleeps.

Marcus: ``Hahae! Father sleeps and doesn't hear you!''

Aemilia: ``Fu**, boy!'' Aemilia is angry. The mother hits the son: slap, slap...

Marcus cries: ``Uhuhu!''

Julius hears him. Father doesn't sleep anymore.

\section[Third scene]{THIRD SCENE}
\emph{Characters: Julius, Aemilia, Julia, Marcus, Quintus.}

Quintus: ``Father comes.''

Aemilia doesn't hear Quintus, because Marcus cries.

Julius sees Quintus and asks him: ``Why does Marcus cry?''

Quintus answers: ``Marcus cries because mother has beat him.''

Julius: ``But why has mother beat Marcus?''

Quintus: ``She has beat Marcus because he is a rude boy. Marcus has beat the little girl!''

Julia: ``Mom! Here is father.'' Aemilia sees Julius.

Aemilia: ``Your son Marcus is rude!''

Julius: ``Fu, boy! A good boy doesn't beat a girl. A boy who beats a little girl is rude!''

Julius, angry, beats the rude boy: slap, slap, slap...

Marcus cries. Quintus is cheerful and laughs. Julia is not cheerful and doesn't laugh either. Why is Julia not cheerful? She is not cheerful, because Marcus cries. Julia is a good girl!

The boy laughs. The girl cries. Who is the boy who laughs? The boy who laughs is Marcus. Who is the girl who cries? The girl who cries is Julia.

Marcus, who has beat the girl, is a rude boy. The girl who Marcus has beat is Julia. Julia calls Aemilia. Aemilia, who Julia has called, the mother of the children is. Aemilia beats the boy. The boy who Aemilia has beat is Marcus.

Who does Quintus call? Quintus calls Julius. Julius, who Quintus has called, the father of the children is. Julius doesn't hear Quintus. Who does Julius hear? Julius hears Marcus. The boy who Julius hears is Marcus.

A girl who sings is cheerful. A girl who cries is not cheerful. A boy who beats a girl is rude!

\section[Grammar]{LATIN GRAMMAR}
\subsection{\emph{Nominative and accusative}}

\begin{enumerate}[(A)]
\item \smallc{Masculine}\\
  Marc\emph{us} ridet. Quintus Marc\emph{um} pulsat.\\
  \emph{-us (-r)} --> \emph{-um}\\
  Ex; Iuli\emph{us}, Iuli\emph{um}; fili\emph{us}, fili\emph{um}; e\emph{um}.
\item \smallc{Feminine}\\
  Iuli\emph{a} cantat. Marcus Iuli\emph{am} pulsat.\\
  \emph{-a} --> \emph{-am}\\
  Ex; puell\emph{a}, puell\emph{am}; parv\emph{a}, parv\emph{am}.
\end{enumerate}

\smallc{Notes;} this declension table with the accusative singular case of masculine and feminine words is limited to \emph{A-}stem and \emph{O-}stem (1st and 2nd declension) sustantives and adjectives.

\subsection{\emph{Verbs}}
Iulia \emph{cantat}. Marcus \emph{ridet}. Iulius \emph{dormit}.

Examples;

\begin{tabular}{| r | r | r |}
  \hline
  \emph{Nominative} & \emph{Accusative} & \emph{Verb}\\
  \hline
  Marcus & Iuliam & pulsat.\\
  Quintus & Marcum & videt.\\
  Iulia & Aemiliam & vocat.\\
  Aemilia & Quintum & interrogat.\\
  ``Quis & me & vocat?''\\
  ``Iulia & te & vocat.''\\
  Marcus & eam & pulsat.\\
  Aemilia & puerum & verberat.\\
  Iulius & eum & audit.\\
  ``Puer probus & parvam puellam & non pulsat!''\\
  Iulius iratus & puerum improbum & verberat.\\
  \hline
\end{tabular}

\smallc{Notes;} only the third person of the present indicative tense is presented here.

\nolinenumbers

\section[Exercises]{EXERCISES}
\subsection*{EXERCISE A}
Cur Marcus Iuliam pulsat? Marcus Iuliam pulsat, quia Iulia cantat. Iulia plorat, quia Marcus eam pulsat. Iulia: ``Mamma! Marcus me pulsat'' Aemilia puellam audit et venit. Mater Quintum videt et eum interrogat: ``Quis me vocat?'' Quintus respondet: ``Iulia te vocat.''

Iulius dormit. Quintus Iulium vocat: ``Pater!'' Marcus ridet quia Iulius non venit. Aemilia Marcum verberat. Iulius venit, quia Marcus plorat. Iulius Aemiliam et Marcum et Quintum et Iuliam videt. Iulius: ``Puer qui parvam puellam pulsat improbus est.'' Iulius puerum improbum verberat. Quem Iulius verberat? Puer quem Iulius verberat est Marcus. Marcus plorat. Puer qui plorat laetus non est. Puella quae cantat laeta est.

\subsection*{EXERCISE B}
Puella cantat: ``Lalla.'' Puella quae cantat est Iulia. Iulia laeta est. Puer improbus puellam pulsat. Puella plorat: ``Uhuhu!'' Puer ridet: ``Hahahae!'' Puer qui ridet est Marcus. Iulia Aemilia vocat: ``Mamma!'' Aemilia venit, et Quintum interrogat: ``Cur Iulia plorat?'' Quintus respondet: ``Iulia plorat, quia Marcus eam pulsat.'' Aemilia: ``Marcus puer probus non est, puer improbus est! Uni est pater?'' Aemilia Iulium non videt. Quintus: ``Pater non hic est.'' Quintus Iulium vocat: ``Pater!'' Iulius Quintum non audit. Cur Iulius non audit? Iulius eum non audit, quia dormit. Marcus plorat, quia Aemilia eum verberat. Quem Aemilia verberat? Aemilia Marcum verberat. Puer quem Aemilia verberat improbus est. Iulia laeta non est neque ridet.

\subsection*{EXERCISE C}
...
