% !TEX root=main.tex

\chapter{MASTER AND SERVANTS}

\linenumbers[1]

\section[First scene]{FIRST SCENE}
\emph{Characters: Julius, Aemilia, Medus.}

The sack of Julius is not small. In his sack there is money. Julius has money in the sack.

Aemilia sees the sack and asks Julius: ``How many coins are in your sack?''

Julius responds: ``A hundred.''

Aemilia: ``What, are there a hundred coins here?''

Julius counts the money: ``One, two, three, four, five, six, seven, eight, nine, ten. What? Ten only?''

Julius counts the money again: ``One, two, three, four... nine, ten.'' The number of coins is not a hundred, but only ten.

Julius: ``What? In my sack there are not a hundred, but only ten coins! Where are the hundred coins? Where are my servants?''

Medus: ``Your servant Medus is here.''

Julius sees his servant Medus, but not Davus. Medus is here. Davus is not present, he is absent. Julius, Aemilia and Medus are present. Davus and the others are absent.

Julius: ``What? Only one servant is present! Where is Davus? Call Davus!''

Medus calls Davus: ``Dave!'' but Davus doesn't hear Medus and doesn't come.

Medus calls Davum again: ``Da-a-vus! Come!''

Davus comes. Now two servants are present.

\section[Second scene]{SECOND SCENE}
\emph{Characters: Julius, Aemilia, Medus, Davus.}
Davus, who doesn't see their master, asks Medus: ``What is the matter, Medus?''

Medus: ``St! The master is here. Salute the master!''

The servant salutes the master: ``Hello, master!''

The master salutes the servant: ``Hello, servant!''

Davus: ``What is the matter, master?''

Julius: ``St! Shut up, servant! Shut up and listen!'' The servant shuts up.

Julius: ``In my sack there are only ten coins. Where are my hundred coins?''

Davus shuts up, and doesn't respond.

Aemilia: ``Respond, Davus! The master asks you.''

Davus answers: ``Your money is not here. Ask Medus!''

Julius asks Medus: ``Where are my coins, Medus?'' Medus responds not a single word.

Julius again asks him: ``Where is my money? Answer, servant!''

Medus accuses Davus: ``Your money is in Davus' sack. Davus has your money.''

Aemilia: ``Listen, Davus! Medus accuses you.''

Davus: ``Who does Medus accuse? Me?''

Julius: ``Silence, Medus! A servant who accuses a servant is rude!''

Julius does not accuse Davus, but asks him: ``Is my money in your sack, Davus?''

Davus: ``Your money is not in my sack, master.''

Julius: ``Where is your sack?''

Davus: ``Here it is. Here is my sack.''

Julius: ``Put your sack on the table!''

Davus puts his sack on the table. Now his sack is on the table. Julius puts his stick on the table. The master's stick is on the table.

Davus: ``See: there are no coins in my sack.''

Julius sees no coins in the sack. In Davus' sack there are no coins. His sack is empty. Davus does not have masters' money.

Julius: ``Oh! Davus is a good servant: he doesn't have my money. Behold your coin, Davus!'' Julius puts one coin in Davus' sack. Now Davus' sack is not empy: in his sack there is one coin. Davus is happy.

Julius: ``Take your sack and leave, good servant!'' Davus takes his sack and leaves.

Medus sees the stick, which is on the table. Medus also leaves! Why does Medus leave? Medus leaves because he has the master's money in his sack!

Davus and Medus are absent.

\section[Third scene]{THIRD SCENE}
\emph{Characters: Julius, Aemilia.}

Julius: ``Davus is a good servant. He doesn't have my money.--But where is my money, Medus? Who has my money?'' Medus doesn't respond.

Julius: ``Where is Medus? Why doesn't he answer?''

Aemilia: ``Medus doesn't answer, because he is absent. No servant is present.''

Julius calls Medus: ``Medus! Come!'' but Medus, who is absent, doesn't hear him nor come.

Julius calls again: ``Me-dus! Come, rude servant!''

Medus doesn't come.

Julius: ``Why doesn't Medus come?''

Aemilia: ``Medus doesn't come, because he has your money! His sack is not empty!'' Aemilia laughs. Julius is angry--he doesn't laugh!

Julius: ``Where is my stick?'' Julius doesn't see his stick, which is on the table.

Aemilia: ``Look, the stick is on the table.''

Julius takes his stick and leaves.

\section[Grammar]{LATIN GRAMMAR}
\subsection{\emph{Vocative}}
\emph{-us --> -e}

Ex; ``Salve, domin\emph{e}''; ``Med\emph{e}! Veni, improb\emph{e} serv\emph{e}!''; Medus Davum vocat: ``Dav\emph{e}!'

\subsection{\emph{Imperative and indicative}}
Dominus: ``Voc\emph{\=a} D\=avum!'' Servus D\=avum voc\emph{at}.

Dominus: ``Tac\emph{\=e}!'' Servus tac\emph{et}.

Dominus: ``Disced\emph{e}, serve!'' Servus disced\emph{it}.

Dominus: ``Aud\emph{\=i}!'' Servus aud\emph{it}.

\begin{enumerate}[(1)]
  \item \emph{-at --> -\=a}
  \item \emph{-et --> -\=e}
  \item \emph{-it --> -e}
  \item \emph{-it --> -\=i}
\end{enumerate}

\section[Exercises]{EXERCISES}
\subsection*{EXERCISE A}

\subsection*{EXERCISE B}

\subsection*{EXERCISE C}
